---
title: "Focus Group Highlights"
author: "Evangeline Reynolds"
date: "5/26/2023"
output: html_document
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE)
```



```{r}
#| label: cars
#| eval: !expr F
#| echo: false
readLines("../easy-geom-recipes-raw-data/ggplot2 extension focus group 1.vtt")
readLines("../easy-geom-recipes-raw-data/Focus Group 2 ggplot2 easy geom recipes .vtt")
```



#### mechanics

> '... going through this was super helpful
because now I like understand the
mechanics of it all.

> 'And so I
don't have any intentions of
like making formal geoms on my
own for anything yet. But it was
like really helpful for
understanding how the whole
system works.'

#### failure w/ previous attempts

> 'So like there's some other layer of
getting into the ggplot
extension world that I [was]
missing.'


#### Step-by-step and including Step 0

> 'So pedagogically, I liked how it
was. I like the general
steps. Like start with ... regular
ggplot and Step 0 just to
have like a baseline. And then
going to each of the steps to
get there and then being able to
compare with the original. Like
as far as like pedagogically,
that was super helpful just as
as an approach to get it
right just so you can have a
goal.  And see how all of
these different primitives and
proto elements and whatever fit together.

# Skepticism

> It was that easy. And I felt
empowered as a result of that.... But you know, like, my problem isn't gonna be that easy.


# Concern regarding missing values

> A place where you might
have an opportunity to do a
little bit of pedagogical
caution [is with] missing
values. So when computing a mean
like ggplot says, by the way, there were three rows I didn't
plot.... So then you
need to explicitly override the
default and that something like
that so that people are not just
blindly putting summaries down
without considering the data
that are being used to make
them.


# Accessibility for students


>  I'm teaching, so I'm
teaching data visualization this
summer online again with my
regular like online classes I've
assigning [the tutorial]... as kind of like an extra
credit thing at the end of
semester to saying like, if
you're interested, go through
this thing and you get 10 bonus
points or something just for the
more advanced students that will
be in the class that will be
interested. 
*But I think it's totally accessible for them.*

# Higher level objectives 

> I'll just add that I I think we, I could definitely use materials
like this and it did raise for
me the points since I didn't
have a lot of experience writing
extensions before like it was
ended up being very
comprehensible to me and so it
kind of made me think that in
like a data visuals, data
visualization classes that we
teach, we probably need to.
There's a balance between
teaching the students to use the
tools that exist right now to
like, just do your analysis and
just do the best with the tools
that are available. But we
probably should include a couple
of weeks on like.
You know, programming and
writing extensions along these
lines, because it's obviously
like very powerful and they need
to have at least some exposure
to it. So you know, a week or
two weeks of materials kind of
like this would be helpful and
would help distinguish them from
like being able to just work
through tutorials on your own
online. Like, if they could
write their own extension,
that's like real value added,
you know, to their organization.
So it it made me think that I
need to think about my data
visualization class A little bit
more as a programming class in
some ways. And I thought that
would be a good.
You know, this is pretty good
material along those lines.

# Relationship to writing functions

> [A useful addition might be] trying to clarify a little bit
more when it's
useful to have
your own custom Geom, as
opposed to your own function.
